Processing and interpreting the personal activity monitoring Data
=================================================================

## Introduction

Loading and preprocessing the data; Read the data in the csv and do the right transformations
```{r}
ActivityData <- read.table("./activity.csv", sep = ",", header = TRUE, na.strings=NA)
ActivityData$date <- as.Date(ActivityData$date)
head(ActivityData, n=10)
```

## What is mean total number of steps taken per day?

We do the following:

1. Summarize the steps per day by using the plyr package. 

2. Plot the histogram of the sums 

3. Calculate and plot the mean and the median
```{r histogram, echo=TRUE}
library(plyr)
df <- ddply(ActivityData, .(date), summarize, sumSteps=sum(steps, na.rm = TRUE))
hist( df$sumSteps, col = "green", breaks=10, xlab="Number of steps per day", main="Histogram of the total number of steps taken each day" )
stepsMean <- mean(df$sumSteps, na.rm = TRUE)
stepsMedian <- median(df$sumSteps, na.rm = TRUE)
abline(v = mean(df$sumSteps, na.rm = TRUE), col = "red", lwd = 4)
abline(v = median(df$sumSteps, na.rm = TRUE), col = "magenta", lwd = 4)
```

The mean value of the total number of steps taken per day is **`r stepsMean`** and the median value of the total number of steps taken per day is **`r stepsMedian`**.

### What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timesSeries, echo=TRUE}
library(plyr)
library(ggplot2)
df2 <- ddply(ActivityData, .(interval), summarize, avgSteps=mean(steps,na.rm = TRUE))
g <- ggplot(df2, aes(x = df2$interval, y = df2$avgSteps))
g + geom_line()+ theme(panel.background = element_rect(colour = "pink"))+ labs(x = "Interval")+ labs(y = "Average Steps per interval")+ labs(title = "Time series plot of the 5-minute interval and the avg num of steps")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
df2 <- ddply(ActivityData, .(interval), summarize, avgSteps=mean(steps,na.rm = TRUE))
maxAvgSteps <- max(df2$avgSteps, na.rm = TRUE)
tmp <- df2[df2$avgSteps == maxAvgSteps,]
maxInterval <- tmp$interval
```
The **`r maxInterval`** 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.

### Imputing missing values
```{r missingValues, echo=TRUE}
totalNumOfNAs<- sum(is.na(ActivityData$steps))
```
There are **`r totalNumOfNAs`** missing values in the dataset (i.e. the total number of rows with NAs).

### Filling in all of the missing values in the dataset with 0
What is mean total number of steps taken per day with the missing values filled in with zeros?
```{r echo=TRUE}
library(plyr)
newActivityData<-ActivityData
newActivityData[is.na(newActivityData)] <- 0
df3 <- ddply(newActivityData, .(date), summarize, sumSteps=sum(steps))
hist( df3$sumSteps, col = "green", breaks=10, xlab="Number of steps per day", main="Histogram of the total number of steps taken each day" )
nstepsMean <- mean(df3$sumSteps, na.rm = TRUE)
nstepsMedian <- median(df3$sumSteps, na.rm = TRUE)
abline(v = mean(df3$sumSteps, na.rm = TRUE), col = "red", lwd = 4)
abline(v = median(df3$sumSteps, na.rm = TRUE), col = "magenta", lwd = 4)
```

The mean value of the total number of steps taken per day is **`r nstepsMean`** and the median value of the total number of steps taken per day is **`r nstepsMedian`**.
